Supporting Information

### 2018-01-25 manuscript.Rmd
added chris's methods bit and a new SI figure

@@

\
The absolute minimum attainable water loss rate is through the cuticle only, which is typically measured on isolated cuticles of the adaxial (non-stomatal) side of the leaf. Other measurements allow for the estimation of water loss through incompletely closed and broken stomata. Of particular interest is the method where leaves are detached from the plant, and their weight loss monitored over time. This method (mass loss of detached leaves, MLD) is interesting because it more closely simulates actual field conditions for plants during severe drought, and includes both surfaces of the leaf - not just the stomata-free side which is typically used for cuticular conductance measurements. A comprehensive compilation of estimates of g~min~ is currently lacking, and is needed not just to parameterize models, but also to study sources of variation in this overlooked plant trait.

\
Previous reviews of the minimum conductance have largely focused on the biology of the plant cuticle, and detailed physiology and anatomy of water transport across cuticles [@riederer2008, @schuster2017; @kerstiens1996; @fernandez2017]. Other work has focused on the role of g~0~ in models of plant water use [e.g. @barnard2013; @lombardozzi2017], without referring to the literature on cuticular conductance. Thus there is a disconnect between the dizzying level of understanding of the plant cuticle [@riederer2008], and the very simple assumptions made when using leaf conductance models in global vegetation models. We aim to improve this connection by synthesizing the state of knowledge. Our goals are to a) compare various definitions of g~min~ quantitively, which have been previously assumed equal, b) demonstrate the need to include a non-zero g~0~ (Eq. 1), for models of water-use efficiency as well as the drought response, c) present a new compilation of minimum conductance measurements, and demonstrate that the large acclimatory potential of this trait to environmental drivers.

@@

\
The minimum conductance of intact leaves is typically measured by mass loss of detached leaves (MLD). In this method, leaves are detached and leaf mass is monitored over time as the leaf dries out. Early work by @hygen1951 showed that after a leaf is detached, initial water loss rates are high, but after a while a constant low rate is achieved. From this minimum transpiration rate (sometimes described as the 'residual transpiration', or 'epidermal transpiration'), the minimum conductance (g~min~) can be estimated using measured vapour pressure deficit [@sinclair1986]. This method most closely resembles conditions that plants may experience during a dry-down or periods of extreme stress.

\
We quantitatively compared each of the different 'minimum' conductance (Table 1) values by synthesizing existing data. To estimate g~0~ (Eq. 1) from regression, we used the @lin2015 database to fit the model of @medlyn2011 using non-linear regression, and data presented by @miner2017 on the linear regression of the @ballberry1987 model. Measurements of conductance of isolated, non-stomatal cuticles (g~cut~) were taken from the compilations by @kerstiens1996 and @schuster2017.  Data on minimum conductance (g~min~) from MLD was compiled from 40 original papers (see SuppInfo 2 for references). For the compilation, we returned to all original papers mentioned in the review by @kerstiens1996, and added many newer sources (see Supporting Information for full details of the compilation). All conductance values were converted to per unit projected surface area. Although it is quite clear that night-time conductance (g~dark~) is much larger than a true 'minimum' conductance (discussed further in Section VI), it is frequently used as an independent measurement of g~0~ [@lombardozzi2017], and thus a comparison of magnitudes is useful. A compilation of night-time conductance (g~dark~) values was taken from @lombardozzi2017, who compiled cuvette (and sometimes sap flow-based) measurements of ca. 150 species during night-time. Actual minimum daytime values were estimated from the @lin2015 database, by taking those data where A~n~ < 2 $mu$ mol m^-2^ s^-1^, and calculating the minimum g~s~ for each of the 52 studies included in that dataset. This estimate was added as a reference as it represents an actual minimum values measured, rather than a fitted or extrapolated g~0~ term.

@@

\
A number of studies have assumed that the g~0~ in the Ball-Berry model equals the night-time conductance (g~dark~) [@barnard2013; @lombardozzi2017], simply because it is a condition when net photosynthesis is zero. However as our literature comparison showed (Fig. 1), g~dark~ is substantially higher compared to any other measure of minimum conductance. The simple idea that g~dark~ is equal to the minimum reached during periods of near-zero PPFD or high temperature during the day can be dismissed based on several observations.

@@

In order to understand variation in g~min~, we have to know more about the role of the cuticle vs. leaky or incompletely closed stomata. A few studies have directly quantified the stomatal component of g~min~. A detailed study of *Hedera helix*  concluded that 35\% of water loss occurred across the stomatal pores (though closed), and 65\% across the non-stomatal part of the cuticle, despite the fact that the stomata presumably covered only a small fraction of the leaf [@santrucek2004]. For ten Sorghum genotypes, @muchow1989 reported a strong positive correlation between stomatal density and g~min~, suggesting leaky stomata contribute substantially to g~min~. For seven conifer species, @brodribb2014 reported that 50 - 94\% of water loss of detached leaves originated from the stomatal side, concluding that stomata must be very leaky. However @santrucek2004 found that the cuticle of the stomatal side (but excluding stomata itself) in *Hedera helix* was many times more more permeable than the non-stomatal side, thus providing another potential explanation for this difference. Either way, water loss rates of intact leaves cannot be simply explained from the permeability of isolated non-stomatal cuticles, and the contribution of leaky stomata, and potentially more permeable cuticle on the abaxial side, likely varies between species.

>>>>

# Notes

@xu1995 lower cuticular transpiration in droughted tomato plants, and increased wax content. Similar response to high salinity, which also reduces water availability.

@sack2003 lower gmin in sun leaves (for 2 out of 4 species), again in the same direction as evaporative demand for sun leaves is frequently higher.

@mengel1989 increased gmin in P. abies after treatment with acidic fog, the epicuticular wax layer showing a 'melted appearance'.

@@

\
The absolute minimum attainable water loss rate is through the cuticle only, which is typically measured on isolated cuticles of the adaxial (non-stomatal) side of the leaf. Other measurements allow for the estimation of water loss through incompletely closed and broken stomata. Of particular interest is the method where leaves are detached from the plant, and their weight loss monitored over time. This method (mass loss of detached leaves, MLD) is interesting because it more closely simulates actual field conditions for plants during severe drought, and includes both surfaces of the leaf - not just the stomata-free side. A comprehensive compilation of estimates of g~min~ is currently lacking, and is needed not just to parameterize models, but also to study sources of variation in this overlooked plant trait.

\
Previous reviews of the minimum conductance have largely focused on the biology of the plant cuticle, and detailed physiology and anatomy of water transport across cuticles [@riederer2008, @schuster2017; @kerstiens1996; @fernandez2017]. Other work has focused on the role of g~0~ in models of plant water use [e.g. @barnard2013; @lombardozzi2017], without referring to the literature on cuticular conductance. Thus there is a disconnect between the dizzying level of understanding of the plant cuticle [@riederer2008], and the very simple assumptions made when using leaf conductance models in global vegetation models. We aim to improve this connection by synthesizing the state of knowledge. Our goals are to a) compare various definitions of g~min~ quantitively, which have been previously assumed equal, b) demonstrate the need to include a non-zero g~0~ (Eq. 1), for models of water-use efficiency as well as the drought response, c) present a new compilation of minimum conductance measurements, and demonstrate the large acclimatory potential of this trait to environmental drivers.

@@

\
The minimum conductance of intact leaves is typically measured by mass loss of detached leaves (MLD). In this method, leaves are detached and leaf mass is monitored over time as the leaf dries out. Early work by @hygen1951 showed that after a leaf is detached, initial water loss rates are high, but after some time a constant low rate is achieved. From this minimum transpiration rate (sometimes described as the 'residual transpiration', or 'epidermal transpiration'), the minimum conductance (g~min~) can be estimated using measured vapour pressure deficit [@sinclair1986]. This method most closely resembles conditions that plants may experience during a dry-down or periods of extreme stress. We do not discuss the methodology of this measurement in detail, but note that care needs to be taken to yield reliable estimates (see SuppInfo for more discussion and illustration).

\
We quantitatively compared each of the different 'minimum' conductance (Table 1) values by synthesizing existing data. To estimate g~0~ (Eq. 1) from regression, we used the @lin2015 database to fit the model of @medlyn2011 using non-linear regression, and data presented by @miner2017 on the linear regression of the @ballberry1987 model. Measurements of conductance of isolated, non-stomatal cuticles (g~cut~) were taken from the compilations by @kerstiens1996 and @schuster2017 (both also include g~min~ data but were not used since they included no metadata). Data on minimum conductance (g~min~) from MLD was compiled from 40 original papers (see SuppInfo 2 for references). For the compilation, we returned to all original papers mentioned in the review by @kerstiens1996, and added many newer sources (see Supporting Information for full details of the compilation). All conductance values were converted to per unit projected surface area. Although it is quite clear that night-time conductance (g~dark~) is much larger than a true 'minimum' conductance (discussed further in Section VI), it is frequently used as an independent measurement of g~0~ [@lombardozzi2017], and thus a comparison of magnitudes is useful. A compilation of night-time conductance (g~dark~) values was taken from @lombardozzi2017, who compiled cuvette (and sometimes sap flow-based) measurements of ca. 150 species during night-time. Actual minimum daytime values were estimated from the @lin2015 database, by taking those data where A~n~ < 2 $mu$ mol m^-2^ s^-1^, and calculating the minimum g~s~ for each of the 52 studies included in that dataset. This estimate was added as a reference as it represents an actual minimum values measured, rather than a fitted or extrapolated g~0~ term.

@@

\
A number of studies have assumed that the g~0~ in the Ball-Berry model equals the night-time conductance (g~dark~) [@barnard2013; @lombardozzi2017], simply because it is a condition when net photosynthesis is zero. However as our literature comparison showed (Fig. 1), g~dark~ is substantially higher compared to any other measure of minimum conductance (except fitted g~0~). The simple idea that g~dark~ is equal to the minimum reached during periods of near-zero PPFD or high temperature during the day can be dismissed based on several observations.

@@

In order to understand variation in g~min~, we have to know more about the role of the cuticle vs. leaky or incompletely closed stomata. A few studies have directly quantified the stomatal component of g~min~. A detailed study of *Hedera helix*  concluded that 35% of water loss occurred across the stomatal pores (though closed), and 65% across the non-stomatal part of the cuticle, despite the fact that the stomata presumably covered only a small fraction of the leaf [@santrucek2004]. For ten Sorghum genotypes, @muchow1989 reported a strong positive correlation between stomatal density and g~min~, suggesting leaky stomata contribute substantially to g~min~. For seven conifer species, @brodribb2014 reported that 50 - 94% of water loss of detached leaves originated from the stomatal side, concluding that stomata must be very leaky. However @santrucek2004 found that the cuticle of the stomatal side (but excluding stomata itself) in *Hedera helix* was many times more more permeable than the non-stomatal side, thus providing another potential explanation for this difference. Either way, water loss rates of intact leaves cannot be simply explained from the permeability of isolated non-stomatal cuticles, and the contribution of leaky stomata, and potentially more permeable cuticle on the abaxial side, likely varies between species.

### 2018-01-25 manuscript_suppinfo.Rmd
added chris's methods bit and a new SI figure

# 2. Methods description for case studies

@@

# 3. Effects of drought stress on g~min~

@@

# 4. Effects of leaf age on g~min~

@@

**Figure S3**. Correlation between g~min~ and leaf age for five studies in the g~min~ database. Leaf age is expressed in years (0 = current year), with the exception of @phillips2010, who measured juvenile and adult foliage on *Eucalyptus*. Citations with species studied are @boyce2000 (*Abies lasiocarpa*, *Picea engelmanii*), @boyce2003 (*Tsuga canadensis*, *Pinus strobus*, *Pinus resinosa*, *Picea rubens*), @heinsoo1998 (*Picea abies*), @jordan2007 (*Agastachys odorata*), @phillips2010 (*Eucalyptus saligna*, *E. sideroxylon*, *E. cladocalyx*).

>>>>

# 2. Methodological issues with measurement of g~min~

For measurements of gmin in detached leaves, the relationship between the change in mass (due to water loss) and time is typically curvilinear and characterised by three phases: 1)the rate of water loss initially increases as stomata temporally open in response to being placed under light; 2) the rate of water loss declines as stomata begin to close in response to increasing leaf water deficit; and 3) once stomata have closed (as much as possible) the rate of water loss is at minimum and is constant over time until leaf relative water content nears zero. In this case, g~min~ is calculated from the linear portion of the curve during phase 3. To achieve this type of response, it is important to ensure that the environmental conditions under which g~min~ is measured are conducive to stomatal closure. From experience, leaves placed under high light and/or high VPD conditions are often unable to effectively close stomata in response to desiccation, possibly as a result of rapid epidermal drying and hydro-passive stomatal opening [@buckley2005]. These conditions lead to a rapid decline in leaf water status and an overestimation of g~min~, as compared to control leaves from the same species.

\
Additionally, sometimes the stomatal response of replicate leaves from the same species can be highly variable, even under the same environmental conditions (see Figure S3). In light of these potential issues, we recommend measurements of g~min~ on detached leaves to be conducted under mild conditions (low light and low VPD), or at least that the measurement conditions are tested for different species and adjusted if necessary. We also note that previous studies have shown that g~min~ decreases substantially with very low relative water content [@gardingen1992]; again it is beneficial to dry the leaf slowly.

```{r goodbadfig, echo=FALSE, dpi=300, fig.width=8, fig.height=4}
figure_goodandbad(goodbadcurves)
```
**Figure S3**. Two example curves for determination of minimum conductance (g~min~) via mass loss of detached leaves (MLD). Leaf weights are expressed as relative water content (RWC). The good quality curve shows a distinct linear portion at the end of the drying curve, from which minimum transpiration can be calculated (using actual leaf weight, and leaf area). The poor quality curve shows a quicker decrease towards the end of the drying curve (> 400min), indicating an *increased* transpiration rate despite the low water content. We speculate that this increase - which we have observed in a number of curves - is due to the epidermis drying so quickly that the stomata are 'pulled open', increasing water loss.

# 3. Methods description for case studies

@@

# 4. Effects of drought stress on g~min~

@@

# 5. Effects of leaf age on g~min~

@@

**Figure S4**. Correlation between g~min~ and leaf age for five studies in the g~min~ database. Leaf age is expressed in years (0 = current year), with the exception of @phillips2010, who measured juvenile and adult foliage on *Eucalyptus*. Citations with species studied are @boyce2000 (*Abies lasiocarpa*, *Picea engelmanii*), @boyce2003 (*Tsuga canadensis*, *Pinus strobus*, *Pinus resinosa*, *Picea rubens*), @heinsoo1998 (*Picea abies*), @jordan2007 (*Agastachys odorata*), @phillips2010 (*Eucalyptus saligna*, *E. sideroxylon*, *E. cladocalyx*).
